---
title: "Minimal hvCpG detection example"
author: "Alice Balard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Minimal hvCpG detection example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(
eval = TRUE,
message = FALSE,
warning = FALSE,
echo = TRUE,
fig.align = "center",
fig.width = 12, # base width in inches
fig.height = 6,
out.width = "100%" # make the figure scale to container width
)

`%>%` <- magrittr::`%>%`
```

## The model in brief

For each CpG, `hyperVarMeth` estimates the **posterior probability that the site is
hypervariable (hv)** across people: `P(Z = 1 | data)`, a value in (0, 1). The model
is hierarchical with three levels:

- **CpG level (Z):** the CpG is hv (Z = 1) with global prior probability `alpha0`,
  or not (Z = 0). There is a single Z shared by all datasets.
- **Dataset level (Z_k):** given Z, each dataset shows the hv pattern with
  probability `p1` (if hv) or slips into looking hv with probability `1 - p0`
  (if not). Datasets are independent *given Z*.
- **Individual level:** given a dataset's state, each person's methylation is a
  draw from a narrow (stable, `sd0`) or wide (hv, `sd1 = lambda * sd0`) Normal.

Because a single CpG's data usually points clearly one way once the dataset
evidences are combined, the per-CpG **posterior** (not a maximum-likelihood
`alpha`) is the meaningful graded output. It depends on the global prior
`alpha0`, which sets the genome-wide baseline rate of hv CpGs.

## 1. Create a mock methylation dataset

```{r}
outdir <- system.file("mock_data", package = "hyperVarMeth")
mock <- hyperVarMeth::create_mock_hvCpG_data(outdir = outdir)
```

This creates a small mock dataset with a few CpGs, samples, and datasets for testing.

## 2. Prepare data

```{r}
# Use system.file() to get installed package data path
data_dir <- system.file("mock_data", package = "hyperVarMeth")

prep <- hyperVarMeth::prepData("mock", dataDir = data_dir)
str(prep)
```

prep contains:

- metadata: sample and dataset information
- medsd_lambdas: dataset-level parameters (`median_sd`, `lambda`)
- cpg_names_all: all CpG names
- h5file: path to HDF5 methylation matrix

```{r}
metadata <- read.table(mock$metadata, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
h5file <- mock$h5file

# Read methylation matrix from HDF5
mat <- rhdf5::h5read(h5file, "matrix", native = TRUE)
rownames(mat) <- mock$cpg_names
colnames(mat) <- metadata$sample

# Compute per-CpG SD across samples
cpg_sd <- apply(mat, 1, sd, na.rm = TRUE)

# Melt matrix for plotting
mat_df <- reshape2::melt(mat, varnames = c("CpG", "Sample"), value.name = "Beta")
mat_df <- merge(mat_df, metadata, by.x = "Sample", by.y = "sample")

ggplot2::ggplot(mat_df, ggplot2::aes(x = Beta, group = dataset, fill = dataset)) +
  ggplot2::geom_histogram(binwidth = 0.05, color = "black") +
  ggplot2::facet_wrap(~ dataset, ncol = 3) +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(legend.position = "none") +
  ggplot2::labs(title = "Beta-value distributions per dataset", x = "Beta", y = "Count")
```

## 3. Run hvCpG detection (batched + parallel)

```{r}
result <- local({
  tmp_dir <- withr::local_tempdir(pattern = "tmp_vignette")

  hyperVarMeth::runAndSave_fast(
    analysis = "mock",
    cpg_names_vec = prep$cpg_names_all,
    resultDir = tmp_dir,
    NCORES = 1,
    p0 = 0.95,          # Pr(dataset looks non-hv | CpG not hv)
    p1 = 0.65,          # Pr(dataset looks hv     | CpG hv)
    alpha0 = 0.05,      # global prior: genome-wide baseline rate of hv CpGs
    dataDir = data_dir,
    batch_size = 100,
    skipsave = FALSE
  )

  result_files <- list.files(tmp_dir, pattern = "results_mock.*\\.rds$")
  if (length(result_files) == 0) stop("No results file generated!")

  # Read the first results file into an R object
  res <- readRDS(file.path(tmp_dir, result_files[1]))

  # Return only the result object; tmp_dir and files deleted on block exit
  res
})
```

This computes the posterior probability of being hv (`post_hv`) for every CpG in
the mock dataset and saves the result to `tmp_dir`. The `alpha0` prior shifts all
posteriors up or down; here we use 0.05 (a plausible genome-wide baseline).

## 4. Inspect results

### Check the first few results
```{r}
head(result)
```

The result is a matrix of CpGs (rows) with a single `post_hv` column: the
posterior probability P(hv | data), between 0 and 1.

```{r}
ggplot2::ggplot(data.frame(post_hv = result[, "post_hv"]),
                ggplot2::aes(x = post_hv)) +
  ggplot2::geom_histogram(bins = 30, colour = "black", fill = "grey70") +
  ggplot2::geom_vline(xintercept = 0.5, linetype = 2) +
  ggplot2::scale_x_continuous(limits = c(0, 1)) +
  ggplot2::labs(x = "Posterior P(hv | data)", y = "CpGs") +
  ggplot2::theme_minimal(base_size = 14)
```

Most CpGs sit near 0 (confidently non-hv); a minority push toward 1 (candidate
hvCpGs). The exact spread depends on `alpha0` and on how separable the mock
stable/hv datasets are.

### Compare a stable, a mid, and a variable CpG

```{r}
post <- result[, "post_hv"]

# lowest, closest-to-0.5, and highest posterior CpGs
sta <- rownames(result)[which.min(post)]
mid <- rownames(result)[which.min(abs(post - 0.5))]
var <- rownames(result)[which.max(post)]

# their posteriors
result[match(c(sta, mid, var), rownames(result)), , drop = FALSE]

sd_df <- mat_df %>%
  dplyr::filter(CpG %in% c(sta, mid, var)) %>%
  dplyr::group_by(dataset, CpG) %>%
  dplyr::summarise(range = max(Beta, na.rm = TRUE) - min(Beta, na.rm = TRUE),
                   sd = sd(Beta, na.rm = TRUE),
                   n = sum(!is.na(Beta)), .groups = "drop")

sd_df$CpG <- factor(sd_df$CpG, levels = c(sta, mid, var),
                    labels = c("stable (low P)", "mid (P~0.5)", "variable (high P)"))

ggplot2::ggplot(sd_df, ggplot2::aes(x = dataset, y = sd, fill = CpG)) +
  ggplot2::geom_col(position = "dodge") +
  ggplot2::coord_cartesian(ylim = c(0, 1)) +
  ggplot2::scale_fill_viridis_d() +
  ggplot2::labs(title = "Per-dataset SD for a stable, mid, and variable CpG",
                x = "Dataset", y = "SD of Beta") +
  ggplot2::theme_minimal(base_size = 14)
```

The high-posterior CpG should show larger per-dataset spread across most datasets;
the low-posterior one should be tight everywhere. A `post_hv` near 0.5 means the
evidence is genuinely ambiguous given the prior.

### Sensitivity to the prior `alpha0` (optional)

Because `post_hv` depends on the global prior, it is worth checking that the
*ranking* of CpGs is stable even as absolute values shift with `alpha0`. Re-run
section 3 with `alpha0 = 0.01` and `alpha0 = 0.1` and compare
`rank(post_hv)` — the ordering should be largely preserved even though the
histogram moves.

## Notes

For real datasets, increase `batch_size` and `NCORES` appropriately.

Use `skipsave = TRUE` if you do not want to write files to disk during testing.

`alpha0` is a genuine modelling input (the genome-wide prior probability that any
CpG is hv), not estimated per CpG — set it from your expected hv rate and report
the value used, since every posterior depends on it.

This vignette demonstrates the minimal workflow using mock data to ensure
everything runs before applying to large-scale datasets.
